Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-09-15 supp/index.Rmd
<edit index>

---
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

>Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918].

Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.

@@

```{r figillust, fig.cap="Illustrated elliptical bifaces from the ancestral Caddo area highlighting oblique parallel flaking used in production."}

@@

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and the associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F. Austin State University.

@@

The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces)), and digitally curated on the Open Science Framework ([DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/)). The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.

>>>>

---
title: "Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices"
author:
    - Robert Z. Selden, Jr.^[Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; Texas Archeological Research Laboratory, The University of Texas at Austin; and Cultural Heritage Department, Jean Monnet University, zselden@sfasu.edu] and John E. Dockall^[Stantec, Inc.]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib]
biblio-style: apalike
link-citations: yes
description: "Pilot study of elliptical biface morphology."
---

@@

>This process of comparison, of recognising in one form a definite permutation or _deformation_ of another, apart altogether from a precise and adequate understanding of the original "type" or standard of comparison, lies within the immediate province of mathematics, and finds its solution in the elementary use of a certain method of the mathematician [@RN7522].

Elliptical bifaces are found at some regularity in Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Elliptical bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This suggests that a shared, and consistent, mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where small bifaces from Jowell Farm were found to be more morphologically diverse than small bifaces from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences that occur in elliptical bifaces.

@@

```{r figillust, fig.cap="Illustrated elliptical bifaces from a, an unknown location; b, Jowell Farm; and c, Belcher Mound highlighting oblique parallel flaking used in production. The specimen from Belcher Mound is a potential preform exhibiting oblique parallel removals consistent with elliptical bifaces from other sites in the region."}

@@

Our thanks to the Caddo Nation of Oklahoma, the Caddo Nation Tribal Council, Tribal Chairman, and Tribal Historic Preservation Office for permission and access to NAGPRA and previously repatriated collections. Our gratitude is also extended to Marybeth Tomka and Lauren Bussiere at the Texas Archeological Research Laboratory for their assistance with access to the bifaces and associated records, to Sergio Ayala and the Gault School of Archaeological Research for access to the UV light, and to Scott Hammerstedt and Debra K. Green at the Oklahoma Archeological Survey for their assistance with records requests. Thanks also to John Harman for access to the DStretch plugin for ImageJ that was useful in the analysis of flake scars, and to Harry J. Shafer, Hiram F. (Pete) Gregory, Christian S. Hoggard, and David K. Thulman for their comments and constructive criticisms on the ongoing analyses of Caddo biface morphology, as well as Emma Sherratt, Kersten Bergstrom, Lauren Butaric, Dean C. Adams, and Michael L. Collyer for their constructive criticisms, general comments, and suggestions throughout the development of this research program.

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was made possible by a Research Support Fund Grant from the Texas Archeological Society. Additional support for this project was provided by the Heritage Research Center at Stephen F. Austin State University.

@@

Reproducibility---the ability to recompute results---and replicability---the chances other experimenters will achieve a consistent result---are two foundational characteristics of successful scientific research [@RN20918]. The analysis code associated with this project can be accessed through this document, is available in the GitHub repository ([https://github.com/seldenlab/elliptical.bifaces](https://github.com/seldenlab/elliptical.bifaces)), and digitally curated on the Open Science Framework ([DOI 10.17605/OSF.IO/PH25W](https://osf.io/ph25w/)). The reproducible nature of this enterprise provides a means for others to critically assess and evaluate the various analytical components [@RN20915;@RN20916;@RN20917], which is a necessary requirement for the production of reliable knowledge.
